---
title: "Prediction of CRISPR guide RNA activity with deep learning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prediction of CRISPR guide RNA activity with deep learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: citations.bib
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```




## Abstract

Deep learning has proven to be one of the state-of-the-art tools in object recognition and natural language processing,
and has been applied to many fields including life sciences.
Deep learning refers to the use of multiple layers of neural network which is one of the traditional machine learning algorithms.
Deep neural networks are able to solve problems that are difficult to solve with conventional machine learning algorithms.
At present, various architectures of deep neural networks have been published,
especially convolutional neural networks (CNN) and recurrent neural networks (RNN) are well-known.
The former is known to be excellent in computer visions such as object recognition and object detection,
while the latter, RNN, is known to be excellent in time-series data analysis and natural language processing.
Considering many Bioconductor users are familiar with genome sequence analysis,
in this workshop, we will focus on RNN,
and introduce the fundamental algorithm of RNN and procedures to build RNN models using R.


### Goals and Objectives

The workshop introduces fundamental concepts of RNN algorithms and long short-term memory (LSTM),
through developing a model to predict CRISPR guide RNA activity with a nucleotide sequence of guide RNA.
By the end of the workshop, participants should be able to:

* understand fundamental concepts of neural network
* understand fundamental concepts of RNN
* develop an deep neural network model with programming language R


### Prerequisties

The workshop will proceed on the assumption that participants have the basic knowledge:

* basic knowledge of R syntax (package installation, reading and writing files, `for` sentence, functions)
* basic knowledge of mathematics (multivariate functions, linear combinations, differentiation, etc.)

Since the workshop focuses on explaining of basic knowledge about deep learning,
it is not intended for those who are familiar with deep learning or who can build models with Python or other programming languages.


## Setup

In this workshop, we mainly use torch [@torch4r] and coro [@coro] packages to perform deep learning,
and use tidyverse package for visualization.
To install these packages, run the following scripts.


```{r, eval = FALSE}
install.packages('tidyverse')
install.packages('coro')
install.packages('torch')
library('torch')
install_torch(timeout = 1200)
```


Then, restart R session and load these packages.


```{r, eval = FALSE}
library('tidyverse')
library('coro')
library('torch')
```


## Dataset Preparation and preprocessing

### Dataset

Dataset used in this workshop can be downloaded from Supplementary Data in @Wang_2019's paper.
Due to time restraints, we will use the dataset that has already been arranged.
The arranged dataset is packaged in Intro2DNN package and also can be downloaded from
[GitHub](https://github.com/jsun/Intro2DNN).
The dataset contains _train.tsv_ and _valid.tsv_ files,
which are the training dataset and validation dataset, respectively.


### Preprocessing

At first, we will show an example of loading and preprocessing of training dataset _train.tsv_.


```{r, eval = FALSE}
train_fpath <- system.file('train.tsv', package = 'Intro2DNN')
crisprcas9_train_dataset <- read.table(train_fpath, header = FALSE, sep = '\t')
head(crisprcas9_train_dataset)
```


As shown above, _train.tsv_ consists of two columns:
the first column is the editing efficiencies of guide RNA,
and the second is the nucleotide sequence of guide RNA.


In this workshop, our goal is to create a model to predict the activity by inputting a nucleotide sequence.
For convenience, we store the activities into `y_train` as labels (response variable)
and the nucleotide sequences into `x_train` as features (explanatory variables).


```{r, eval = FALSE}
x_train <- crisprcas9_train_dataset[, -1]
y_train <- crisprcas9_train_dataset[, 1]
```


Since most machine learning packages including torch package require numeric inputs,
it is necessary to convert a string (i.e., nucleotide sequence) into numeric values.
Based on the requirements of torch package,
we convert A, T, C, and G into integers 1, 2, 3, and 4, respectively.
Here, as an example,
we try to covert the first sequence of `x_train` with `chartr` and `strsplit` functions.


```{r, eval = FALSE}
chartr('ATCG', '1234', x_train[1])
as.integer(unlist(strsplit(chartr('ATCG', '1234', x_train[1]), '')))
```


We can see that the 21 nucleotides was successfully converted into 21 integers
with `chartr` and `strsplit` functions.
Next, we use the two functions to convert all sequences in `x_train` object.


```{r, eval = FALSE}
x_train <- matrix(as.integer(unlist(strsplit(chartr('ATCG', '1234', x_train), ''))),
                  ncol = 21, byrow = TRUE)
head(x_train)
```


With the above steps, we successfully loaded data from a file into R object,
and converted all sequences into numeric values.
Unfortunately, torch package cannot directly handle this type of R object.
It is required to create a torch-readable class (a collection of datasets and functions)
following torch requirements.


```{r, eval = FALSE}
set_dataset <- dataset(
    'guidRNA_dataset',
    
    initialize = function(x, y) {
        self$x <- x
        self$y <- y
    },
    
    .getitem = function(i) {
        x_tensor <- torch_tensor(as.integer(self$x[i, ]))
        y_tensor <- torch_tensor(as.numeric(self$y[i]))
        list(x = x_tensor, y = y_tensor)
    },
    
    .length = function() {
        nrow(self$x)
    }
)
```


Then, we assign the training data `x_train` and `y_train` to `set_dataset` class;
and we pass `set_dataset` class object to `dataloader` to manage datasets during model training.


```{r, eval = FALSE}
dataset_train <- set_dataset(x_train, y_train)
dataloader_train <- dataloader(dataset_train, batch_size = 1024, shuffle = TRUE)
```



## Modeling

### Model Architecture

The first step of model construction with torch package is to design architecture of neural networks.
Here, we design a neural network with multiple layers that receives 21 integers (features) and outputs one real value (label).
Specifically, we design an architecture with three main parts:
(i) the first part is composed of an embedding layer that converts 21 integer values to 64 dimension values,
(ii) the second part is composed of an LSTM layer which receives 64 values and outputs 256 features,
and (iii) the third part is composed of three fully connected layers
which receives 256 values and outputs one real value.


To design the network architecture, we follow the definitions of torch to create a class
with `initialize` and `forward` functions.
The `initialize` function declares the components of the network,
and the `forward` function defines the order in which the components are connected.


```{r, eval = FALSE}
GenomicNet <- nn_module(
    "GenomicNet",
  
    initialize = function() {
        self$embedding <- nn_embedding(num_embeddings = 4, embedding_dim = 64)
        self$lstm <- nn_lstm(input_size = 64, hidden_size = 256, batch_first = TRUE)
        self$dropout <- nn_dropout(p = 0.5)
        self$fc1 <- nn_linear(in_features = 256, out_features = 512)
        self$fc2 <- nn_linear(in_features = 512, out_features = 512)
        self$fc3 <- nn_linear(in_features = 512, out_features = 1)
    },
  
    forward = function(x) {
        x <- self$embedding(x)
        x <- self$lstm(x)
        x <- x[[2]][[1]][1, , ]
        x <- nnf_relu(x)
        x <- self$fc1(x)
        x <- self$dropout(x)
        x <- nnf_relu(x)
        x <- self$fc2(x)
        x <- self$dropout(x)
        x <- nnf_relu(x)
        x <- self$fc3(x)
  }
)
```



### Model Training


In this subsection,
we will create an instance from the model architecture and assign the dataset for model training.
Here is an example for creating an instance from the `GenomicNet` class.


```{r, eval = FALSE}
model <- GenomicNet()
```


To train the model, we specify a training algorithm and a loss function in advance.
Since our goal is to predict a single real number, we will solve it as a regression problem.
Here we will use the mean squared error (MSE) as the loss function.
In addition, we will use Adam's algorithm to optimize the model,
which is one of the most popular algorithms in most situations.


```{r, eval = FALSE}
criterion <- nnf_mse_loss
optimizer <- optim_adam(model$parameters)
```


Next, we send the prepared datasets and the model to a device (CPU or GPU) for training.
Here, we use `for` statement to train 5 epochs with the same dataset.
At each epoch, we train the model with each of the minibatches defined by `dataloader`.


```{r, eval = FALSE}
model$to(device = 'cpu')
model$train()

loss_train <- c()

for (epoch in 1:5) {
    loss_running <- 0
    n_train_samples <- 0
  
    # loop for minibatches
    coro::loop(for (b in dataloader_train) {
        optimizer$zero_grad()
        output <- model(b$x$to(device = 'cpu'))
        loss <- criterion(output, b$y$to(device = 'cpu'))
        loss$backward()
        optimizer$step()
        
        loss_running <- loss_running + loss$item() * nrow(b$x)
        n_train_samples <- n_train_samples + nrow(b$x)
    })
    
    loss_train <- c(loss_train, loss_running / n_train_samples)
    cat(sprintf("epoch %d  loss: %3f\n", epoch, loss_running / n_train_samples))
}
```


The training loss decreased during training.
Further, even at the 5th epoch, the downward trend of the training loss has not weakened.
Thus, it is expected that the training loss can be further reduced by setting a larger number of epochs.


```{r, eval = FALSE}
data.frame(epoch = 1:length(loss_train), loss = loss_train) %>%
    ggplot(aes(x = epoch, y = loss)) +
    geom_line()
```


Let us train the model more 5 epochs.


```{r, eval = FALSE}
for (epoch in 6:10) {
    loss_running <- 0
    n_train_samples <- 0

    coro::loop(for (b in dataloader_train) {
        optimizer$zero_grad()
        output <- model(b$x$to(device = 'cpu'))
        loss <- criterion(output, b$y$to(device = 'cpu'))
        loss$backward()
        optimizer$step()
        
        loss_running <- loss_running + loss$item() * nrow(b$x)
        n_train_samples <- n_train_samples + nrow(b$x)
    })
    
    loss_train <- c(loss_train, loss_running / n_train_samples)
    cat(sprintf("epoch %d  loss: %3f\n", epoch, loss_running / n_train_samples))
}


data.frame(epoch = 1:length(loss_train), loss = loss_train) %>%
    ggplot(aes(x = epoch, y = loss)) +
    geom_line()
```



### Model Validation


Here we use the validation dataset to validate the model performance.
The procedures of data preprocessing for validation is the same as that for training.


```{r, eval = FALSE}
valid_fpath <- system.file('valid.tsv', package = 'Intro2DNN')
crisprcas9_valid_dataset <- read.table(valid_fpath, header = FALSE, sep = '\t')

x_valid <- crisprcas9_valid_dataset[, -1]
y_valid <- crisprcas9_valid_dataset[, 1]

x_valid <- matrix(as.integer(unlist(strsplit(chartr('ATCG', '1234', x_valid), ''))),
                  ncol = 21, byrow = TRUE)

dataset_valid <- set_dataset(x_valid, y_valid)
dataloader_valid <- dataloader(dataset_valid, batch_size = 1024, shuffle = FALSE)
```


Then, same as the training steps, we use `for` statement to assign the validation dataset to the model,
and retrieve the prediction results.
Note that, switching the model to validation mode (evaluation mode) enables to improve the calculation speed during validation.


```{r, eval = FALSE}
model$eval()

y_true <- y_valid
y_pred <- c()

loss_valid <- 0
n_valid_samples  <- 0

coro::loop(for (b in dataloader_valid) {
    output <- model(b$x$to(device = 'cpu'))
    y_pred <- c(y_pred, as.numeric(output))
    loss <- criterion(output, b$y$to(device = 'cpu'))
    
    loss_valid <- loss_valid + loss$item() * nrow(b$x)
    n_valid_samples <- n_valid_samples + nrow(b$x)
})

loss_valid <- loss_valid / n_valid_samples
loss_valid
```


Then we plot a scatter chart to visualize the correlation between predicted values and the true values.


```{r, eval = FALSE}
data.frame(label = y_true, predicted = y_pred) %>%
    ggplot(aes(x = label, y = predicted)) +
    geom_point() +
    coord_fixed() +
    xlim(0, 1) + ylim(0, 1)
```


The mean squared error (MSE) was calculated during the validation step, and stored in `loss_valid` object.
In addition, including MSE, the evaluation metrics such as Spearman's rank correlation coefficient
also can be calculated following the definition.


```{r, eval = FALSE}
# MSE
sum((y_true - y_pred)^2) / length(y_true)

# correlation
cor(y_true, y_pred, method = c('spearman'))
```


### Inference

In this subsection, we show an example to infer activities of guide RNA with the trained model.
To perform inference, the scripts used for validation also can be used as-is.
Here, we show another method to infer activity with a single nucleotide sequence of guide RNA.

```{r, eval = FALSE}
x <- 'GAGTGATGATGGTCTGCACAC'
x <- matrix(as.integer(unlist(strsplit(chartr('ATCG', '1234', x), ''))),
                  ncol = 21, byrow = TRUE)
x

x_tensor <- torch_tensor(x)
x_tensor

y <- model(x_tensor$to(device = 'cpu'))
y <- y$item()
y
```



## Saving and Loading

The trained model can be saved with `torch_save` function.
Note that models saved with the standard `save` function will be environment-dependent,
and results in that you are not able to call the model in other environments (computers).


```{r, eval = FALSE}
torch_save(model, 'my_model.pth')
```


The model can be loaded with `torch_load` function from a file.
Models loaded by `torch_load` function can be used for inference or retraining.


```{r, eval = FALSE}
mymodel <- torch_load('my_model.pth')
mymodel$eval()

x <- 'GAGTGATGATGGTCTGCACAC'
x <- matrix(as.integer(unlist(strsplit(chartr('ATCG', '1234', x), ''))),
                  ncol = 21, byrow = TRUE)
x_tensor <- torch_tensor(x)
x_tensor
y <- mymodel(x_tensor$to(device = 'cpu'))
y <- y$item()
y
```



## References {-}

<div id="refs"></div>



